### 2014-10-31 paper/include/background.tex
corrections

DRAM is hierarchically organized as in \reffig{fig:dram_orga}. At the top of the hierarchy are the \textit{channels} that are connected to one or more \textit{ranks}, each composed of multiple storage arrays called \textit{banks}. Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently. All ranks on one channel can operate in parallel, but are constrained by the channel bandwidth, which is shared among the ranks. Further, the banks within the ranks operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks.

@@

To read a row, the word line connected to the row's transistors must first be activated. When the word line is activated, the transistors connect the capacitors to the bit lines. When the capacitors connect to the bit lines, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively. For this to work, the bit lines have to be precharged to $V_{DD}/2$. The sense amplifiers then detect the change of voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}/2$ or $0\:V$ by the sense amplifiers depending on the former voltage detected. Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.

@@

As mentioned, the capacitors leak charge over time and have to be recharged in order to preserve the data. The refresh is performed by opening a row to let the sense amplifiers magnify the detected voltage change and thus recharge the cells to either $V_{DD}/2$ or $0\:V$. This follows the same procedure as a read operation and thus is the row always refreshed when a read is performed. % Is it clear that the row is refreshed during a write?

>>>>

DRAM is hierarchically organized as in \reffig{fig:dram_orga}. At the top of the hierarchy are the \textit{channels} that are connected to one or more \textit{ranks}, each composed of multiple storage arrays called \textit{banks}. Each channel has individual address, data, and command buses that makes it possible for the channels to operate concurrently. All ranks on one channel can operate in parallel, but are constrained by the channel bandwidth, which is shared among the ranks. Further, the banks within the ranks operate in parallel, but are constrained by both the shared channel bandwidth and resources shared among the banks, such as power.

@@

To read a row, the word line connected to the row's transistors must first be activated. When the word line is activated, the transistors connect the capacitors to the bit lines. Then, charge will either flow from or to the capacitors depending on whether they were charged or not, respectively. For this to work, the bit lines have to be precharged to $V_{DD}/2$. The sense amplifiers then detect the change in voltage, which makes it possible to interpret the stored data. The charge stored on the capacitors has, as mentioned, left the capacitor and a read operation is thus destructive. To maintain the data the capacitors are recharged by the bit lines, which are driven to either $V_{DD}$ or $0\:V$ by the sense amplifiers depending on the former voltage detected. Finally, when another row on the bank is accessed, the open row has to be closed, which disconnects the capacitors from the bit lines.

@@

As mentioned, the capacitors leak charge over time and have to be refreshed in order to preserve the data. The refresh is performed by opening a row to let the sense amplifiers magnify the detected voltage change and thus recharging the cells to either $V_{DD}$ or $0\:V$. This follows the same procedure as a read operation and thus is the row always refreshed when a read is performed. % Is it clear that the row is refreshed during a write?

### 2014-10-31 paper/include/introduction.tex
corrections

DRAM is the most commonly used main memory technology in modern computer systems and represent a growing portion of system power consumption \cite{exascale}. This portion will continue to increase as we strive for higher and higher memory capacity. A large part of the DRAM power consumption originates from refreshing all DRAM memory-cells. This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time. To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but stacked design have trouble dissipating heat. The higher temperature increases leakage, and the DRAM will have to be refreshed more often.

There are two main ways to decrease power consumption in DRAM; minimizing unnecessary operations and scheduling the needed ones smarter. As such, one straightforward way to mitigate DRAM power consumption is to reduce the number of unnecessary refresh operations. Here we survey some proposed approaches to avoid or delaying refresh. Smarter scheduling, while promising interesting, is beyond the scope of this paper.

>>>>

DRAM is the most commonly used main memory technology in modern computer systems and represent a growing portion of system power consumption \cite{exascale}. This portion will continue to increase as we strive for higher and higher memory capacity. A large part of the DRAM power consumption originates from refreshing all DRAM memory-cells. This refreshing has to be done as DRAM's store data as a charge in a capacitor, and these charges leak over time. To increase DRAM performance and lower access latencies, 3D stacked DRAM has been proposed, but a stacked design have trouble dissipating heat. The higher temperature increases leakage, and the DRAM will have to be refreshed more often.

There are two main ways to decrease power consumption in DRAM; minimizing unnecessary operations and scheduling the needed ones smarter. As such, one straightforward way to mitigate DRAM power consumption is to reduce the number of unnecessary refresh operations. Here we survey some proposed techniques to avoid or delaying refresh. Smarter scheduling, while promising interesting, is beyond the scope of this paper.

### 2014-10-31 paper/include/reducing_refresh_power.tex
disc text
corrections

\section{Reducing refresh power} 
\label{sec:red}
There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on with sort of information the scheduling decisions are based upon. The four types of information are: access recency, cell retention time, error tolerance of the data, and line validity. This categorization follows the taxonomy proposed by Cui et al. \cite{dtail}. These four approaches are explained further in \ref{sec:red:app}, whereas techniques based on the approaches are summarized in \ref{sec:red:tech}.

@@

\subsubsection*{\textbf{Retention Time (R)}}
The need for refresh is as mentioned caused by leakage in the memory cells. Retention time is a measurement of the time it takes before a cell loose its data if not refreshed (recharged). Variations in the manufacturing process causes a exponentially distributed fluctuation \cite{katayama} in the retention time of the DRAM cells on the same chip. To accommodate for the cells with the lowest retention time manufacturers produce chips with average retention times much higher than possible minimum. This gives room for optimization using variable refresh rate for different cells. The retention time is also highly sensitive to temperature variations and becomes drastically lower at higher temperatures. For techniques using this approach the biggest advantage is that the improvements are independent of workload, except for a potential rise in temperature which would lower the retention time.

\subsubsection*{\textbf{Data Tolerance (T)}}
In the most approaches all data loss is seen as intolerable. In this approach data is categorized depending on how critical it is and e.g. approximate computations, games, or media processing can tolerate some bit errors while still producing a acceptable result. This yields that cells containing data from such applications may be refreshed more seldom. Using this approach the energy and performance savings achieved is highly depending on what kind of applications is most commonly run on the system.

\subsubsection*{\textbf{Validity (V)}}
In a system that does not utilize all the available memory space it is unnecessary to refresh parts not used. Therefore, cells containing data that are not valid does not need to be refreshed. This approach is highly sensitive to memory utilization of the system and its effects on energy consumption and performance improvements will be diminishing on systems with high workload.

>>>>

\section{Reducing refresh power} 
\label{sec:red}
There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on which sort of information the scheduling decisions are based upon. The four types of information are: access recency, cell retention time, error tolerance of the data, and line validity. This categorization follows the taxonomy proposed by Cui et al. \cite{dtail}. These four approaches are explained further in \ref{sec:red:app}, whereas techniques based on the approaches are summarized in \ref{sec:red:tech}.

@@

\subsubsection*{\textbf{Retention Time (R)}}
The need for refresh is as mentioned caused by leakage in the memory cells. Retention time is a measurement of the time it takes before a cell loose its data if not refreshed (recharged). Variations in the manufacturing process causes a exponentially distributed fluctuation \cite{katayama} in the retention time of the DRAM cells. To accommodate for the cells with the lowest retention time, manufacturers produce chips with average retention times much higher than possible minimum. This gives room for optimization using variable refresh rate for different cells. The retention time is also highly sensitive to temperature variations and becomes drastically lower at higher temperatures. For techniques using this approach the biggest advantage is that the improvements are independent of workload, except for a potential rise in temperature which would lower the retention time.

\subsubsection*{\textbf{Data Tolerance (T)}}
In the other approaches all data loss is seen as intolerable. In this approach data is categorized depending on how critical it is, which yields that cells containing data with a lower criticality may be refreshed more seldom. Applications such as approximate computations, games, or media processing can tolerate some bit errors while still producing a acceptable result. Using this approach, the energy and performance savings achieved is highly depending on what kind of applications most commonly are run on the system.

\subsubsection*{\textbf{Validity (V)}}
In a system that does not utilize all the available memory space it is unnecessary to refresh parts not used. Therefore, cells that do not keep valid data do not need to be refreshed. The allocation and deallocation of the cells have to be tracked to get the information needed for implementation. This approach is highly sensitive to memory utilization of the system and its effects on energy consumption and performance improvements will be diminishing on systems with high workload.
